---
title: "Cyclistic Bike Share Analysis"
output:
  html_document:
    df_print: paged
---
### Load library packages
```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
library(ggplot2)
library(scales)
library(ggplot2)
options(scipen = 999)
```

### Read the 12 CSV files
```{r}
apr_2021 <- read_csv("../input/bike-trips-data/202104-divvy-tripdata.csv")
may_2021 <- read_csv("../input/bike-trips-data/202105-divvy-tripdata.csv")
jun_2021 <- read_csv("../input/bike-trips-data/202106-divvy-tripdata.csv")
jul_2021 <- read_csv("../input/bike-trips-data/202107-divvy-tripdata.csv")
aug_2021 <- read_csv("../input/bike-trips-data/202108-divvy-tripdata.csv")
sep_2021 <- read_csv("../input/bike-trips-data/202109-divvy-tripdata.csv")
oct_2021 <- read_csv("../input/bike-trips-data/202110-divvy-tripdata.csv")
nov_2021 <- read_csv("../input/bike-trips-data/202111-divvy-tripdata.csv")
dec_2021 <- read_csv("../input/bike-trips-data/202112-divvy-tripdata.csv")
jan_2022 <- read_csv("../input/bike-trips-data/202201-divvy-tripdata.csv")
feb_2022 <- read_csv("../input/bike-trips-data/202202-divvy-tripdata.csv")
mar_2022 <- read_csv("../input/bike-trips-data/202203-divvy-tripdata.csv")
```

### Combine datasets
```{r}
bike_tripdata <- rbind(apr_2021, may_2021, jun_2021, jul_2021, aug_2021, sep_2021, oct_2021, nov_2021, dec_2021, jan_2022, feb_2022, mar_2022)
```

### Check data type and data structure
```{r}
glimpse(bike_tripdata)
```

```{r}
skimr::skim(bike_tripdata)
```

## Data cleaning and manipulation

### Remove all NA values 
```{r}
bike_tripdata <- na.omit(bike_tripdata)
```

### Remove rows with start time later than end time
```{r}
bike_tripdata <- bike_tripdata[! (bike_tripdata$started_at > bike_tripdata$ended_at),]
```

### Add columns (ride duration / day of week / month / hour) for further analysis 
```{r}
bike_tripdata <- bike_tripdata %>% 
  mutate(duration = as.numeric(difftime(ended_at, started_at, units = "mins"))) %>% 
  mutate(day_of_week = format(started_at, format ="%A")) %>% 
  mutate(month = format(started_at, format = "%B_%Y")) %>% 
  mutate(hour = as.numeric(difftime(ended_at, started_at, units = "hours"))) %>% 
  glimpse()
```

### Remove ride duration that is "0"
```{r}
bike_tripdata <- bike_tripdata[! (bike_tripdata$duration <= 0),]
```

### Order day_of_week and month
```{r}
bike_tripdata$day_of_week <- ordered(bike_tripdata$day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday","Saturday", "Sunday"))
```

```{r}
bike_tripdata$month <- ordered(bike_tripdata$month, levels = c("April_2021", "May_2021", "June_2021", "July_2021", "August_2021", "September_2021", "October_2021", "November_2021", "December_2021", "January_2022", "February_2022", "March_2022"))
```

### Analyze the min / max / avg / median ride duration of member and casual riders
```{r}
bike_tripdata %>% 
  group_by(member_casual) %>% 
    summarize(min_duration = min(duration), max_duration = max(duration), 
              avg_duration = mean(duration), median_duration = median(duration))
```

### Analyze the mean duration of member and casual riders by day of week
```{r}
bike_tripdata %>% 
  group_by(member_casual, day_of_week) %>% 
    summarize(bike_count = n(), avg_duration = mean(duration))
```

### Total Number of Rides for Each User Type
```{r}
bike_tripdata %>% 
  ggplot(aes(x=member_casual)) +
  geom_bar(fill="#09a3b8") +
  labs(
    title = "Total Number of Rides for Each User Type",
    subtitle = "(April 2021 to March 2022)",
    x = "Type of User",
    y = "Number of rides"
  ) +
  scale_y_continuous(labels = label_number(suffix = " M", scale = 1e-6)) +
  geom_text(stat = "count", aes(label = ..count..), position = position_stack(vjust = 0.5))
```

### Bike Preference by User Type
```{r}
bike_tripdata %>% 
  group_by(rideable_type,member_casual) %>% 
  summarize(bike_count = n()) %>% 
  ggplot(aes(x = member_casual,
             y = bike_count,
             fill = rideable_type, 
             )) +
  geom_bar(stat = "identity") +
  labs(
    title = "Bike Preference by User Typer",
    subtitle = "(April 2021 to March 2022)",
    x = "Type of User",
    y = "Bike Count",
    fill = "Bike Type"
  ) +
  geom_text(aes(label = bike_count), position = position_stack(vjust = 0.5))
```

### Number of Rides per Month for Each User Type
```{r}
bike_tripdata %>% 
  group_by(member_casual, month) %>% 
  summarize(bike_count = n()) %>% 
  ggplot(aes(x = month,
             y = bike_count,
             fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    title = "Number of Rides per Month for Each User Type",
    subtitle = "(April 2021 to March 2022)",
    x = "Month_Year",
    y = "Bike Count",
    fill = "User Type"
  ) +
   coord_flip()
```

### Number of Rides per Day of Week for Each User Type
```{r}
bike_tripdata %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(bike_count = n()) %>% 
  ggplot(aes(x = day_of_week,
             y = bike_count,
             fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    title = "Number of Rides per Day of Week for Each User Type",
    subtitle = "(April 2021 to March 2022)",
    x = "Day of Week",
    y = "Bike Count",
    fill = "User Type"
  ) +
  coord_flip()
```

### Total Bike Ride Duration in Hours for Each User Type
```{r}
bike_tripdata %>% 
  select(member_casual, hour) %>% 
  group_by(member_casual) %>% 
  summarize(total_duration = round(sum(hour),2)) %>% 
  ggplot(aes(x = member_casual,
             y = total_duration)) +
  geom_bar(stat = "identity",
           fill = "#FF8000") +
  labs(
    title = "Total Bike Ride Duration in Hours for Each User Type",
    subtitle = "(April 2021 to March 2022)",
    x = "Type of User",
    y = "Total Duration (Hours)"
  ) +
  geom_text(aes(label = total_duration), position = position_stack(vjust = 0.5))
```

### Average Bike Ride Duration by Month for Each User Type
```{r}
bike_tripdata %>% 
  group_by(member_casual, month) %>% 
  summarize(average_duration = round(mean(duration),2)) %>% 
  ggplot(aes(x = month,
             y = average_duration,
             fill = member_casual)) + 
  geom_bar(stat ="identity") +
  coord_flip() +
  facet_wrap(member_casual~.) +
  labs(
    title = "Average Bike Ride Duration by Month for Each User Type",
    subtitle = "(April 2021 to March 2022)",
    x = "Month_Year",
    y = "Duration (minutes)",
    fill = "User Type"
  ) +
  coord_flip() +
    geom_text(aes(label = average_duration), position = position_fill(), hjust = -2.15)
```

### Average Bike Ride Duration by Day of Week for Each User Type
```{r}
bike_tripdata %>% 
  group_by(member_casual, day_of_week) %>% 
  summarize(average_duration = round(mean(duration),2)) %>% 
  ggplot(aes(x = day_of_week,
             y = average_duration,
             fill = member_casual)) + 
  geom_bar(stat ="identity") +
  coord_flip() +
  facet_wrap(member_casual~.) +
  labs(
    title = "Average Bike Ride Duration by Day of Week for Each User Type",
    subtitle = "(April 2021 to March 2022)",
    x = "Day Of Week",
    y = "Duration (minutes)",
    fill = "User Type"
  ) +
  coord_flip() +
    geom_text(aes(label = average_duration), position = position_fill(), hjust = -2.2)
```

